---
title: "Effect of climate change on Bistorta vivipara traits on Svalbard"
format: html
editor: visual
bibliography: references.bib
csl: nordic-journal-of-botany.csl
execute:
  echo: false
---

```{r}
#| label: load-packages
#| message: false

library(here)      # for help finding files
library(tidyverse) # for ggplot2, dplyr, readr etc

# default plot theme
theme_set(theme_bw())

```

# Introduction

Climate change is expected to have large impacts on both marine [@perry2005] and terrestrial ecosystems in Norway.

# Methods

## Field methods

The experiment was carried out in the high Arctic, in Endalen (78°11'N, 15°45'E), approximately four kilometres east of Longyearbyen, Svalbard, at 80-90 m elevation. The experiment was established in 2002 in three different habitats which are located in the south--southeast--facing hillside of the valley. The habitats differ in vegetation composition and the time of snowmelt (and hence the length of the growing season). The habitats include a relatively dry Dryas heath with thin snow cover (ca. 10 cm) and early snowmelt, a mesic Cassiope heath habitat with intermediate snow depth and melting date, and a moist snowbed community with deep snow (\> 100 cm) and late snowmelt.

In 2001, ten plots (75x75 cm) were selected in each of the three habitats. Half of the plots were randomly assigned to the warming treatment in 2002 (open-top chambers, OTC) and the other half to control (CTL).

## Data Analysis

To test the effect of warming on plant height, a linear regression with a normal error distribution was used. Plant height was the reponse variable and warming treatment the predictor. Data were analysed in R version `r paste(R.version$major, R.version$minor, sep = ".")` [@R], using tidyverse for data manipulation and plots [@tidyverse]. The trait data are archived at [OSF](https://osf.io/smbqh).

# Results

```{r}
#| label: load-data
#| message: false

# import data
traits <- read_delim(
    file = here("data/PFTC4_Svalbard_2018_ITEX_Traits.csv")) 

# filter bistorta species and plant height
bistorta <- traits |>
  filter(Taxon == "bistorta vivipara",
         Trait == "Plant_Height_cm")

```

```{r}
#| label: model

# run a linear model
mod_height <- lm(Value ~ Treatment, data = bistorta)

```

In total, we measured the height of 77 *Bistorta vivipera* plants across the different habitats. The *Bistorta vivipera* in the OTC plots is 1.2 cm ± 0.39 taller than in the control plots (@fig-height-effect).

```{r}
#| label: fig-height-effect
#| fig-cap: Mean plant height in control (CTL) and warm (OTC) plots.
#| fig-width: 5.2

ggplot(bistorta, aes(x = Treatment, y = Value)) +
  geom_boxplot(fill = c("grey80", "red")) +
  labs(x = "Treatment", y = "Plant height (cm)")
```
